---
title: "Two Binary Co-Primary Endpoints (Asymptotic Methods)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two Binary Co-Primary Endpoints (Asymptotic Methods)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates sample size calculation for clinical trials with two co-primary binary endpoints using asymptotic normal approximation methods. The methodology is based on Sozu et al. (2010).

```{r setup, message=FALSE, warning=FALSE}
library(twoCoprimary)
library(dplyr)
library(tidyr)
library(knitr)
```

## Background

### Clinical Context

Binary co-primary endpoints are common in:

- **Migraine trials**: Relief from headache (yes/no) + Relief from the most bothersome migraine-related symptom (yes/no)

- **Psoriasis trials**: At least 75% improvement in the score on the psoriasis area-and-severity index (PASI) (yes/no) + Score on the physician's global assessment of 0 or 1 (yes/no)

- **Primary myelofibrosis trials**: Clinically relevant complete haematological response (yes/no) + Lack of progression of clinical symptoms (yes/no)

### When to Use Asymptotic Methods

Asymptotic (normal approximation) methods are appropriate when:

- **Sample sizes are large** (typically $N > 200$)

- **Probabilities are not extreme** ($0.10 < p < 0.90$)

- **Computational efficiency is important**

For small to medium sample sizes or extreme probabilities, use **exact methods** based on Homma and Yoshida (2025), which can be computed using exact approaches without any approximations (see `vignette("two-binary-endpoints-exact")`).

## Statistical Framework

### Model and Assumptions

Consider a two-arm parallel-group superiority trial comparing treatment (group 1) with control (group 2). Let $n_{1}$ and $n_{2}$ denote the sample sizes in the two groups, with allocation ratio $r = n_{1}/n_{2}$ (i.e., total sample size is $N=n_{1}+n_{2}$).

For subject $i$ ($i = 1, \ldots, n_j$) in group $j$ ($j = 1$: treatment, $j = 2$: control), we observe two binary **outcomes**:

$$X_{i,j,k} \in \{0, 1\}, \quad k = 1, 2$$

where $X_{i,j,k} = 1$ indicates success and $X_{i,j,k} = 0$ indicates failure for outcome $k$.

**Marginal probabilities**: Let $p_{j,k}$ denote the success probability for outcome $k$ in group $j$:

$$p_{j,k} = {\rm P}(X_{i,j,k} = 1), \quad k = 1, 2$$

For details on the joint distribution of $(X_{i,j,1}, X_{i,j,2})$, see Homma and Yoshida (2025) Section 2.1 or Sozu et al. (2010).

### Correlation Structure

The correlation $\rho_{j}$ between the two binary outcomes in group $j$ is defined as (Homma and Yoshida, 2025):

$$\rho_j = {\rm Cor}(X_{i,j,1}, X_{i,j,2}) = \frac{\phi_j - p_{j,1}p_{j,2}}{\sqrt{p_{j,1}(1-p_{j,1})p_{j,2}(1-p_{j,2})}}$$

where $\phi_j = {\rm P}(X_{i,j,1} = 1, X_{i,j,2} = 1)$ is the joint probability that both outcomes are successful.

**Practical interpretation**: $\rho_{j} > 0$ means subjects who succeed on outcome 1 are more likely to succeed on outcome 2.

**Valid correlation range**: Because $0 < p_{j,k} < 1$, the correlation $\rho_{j}$ is not free to range over $[-1, 1]$, but is bounded by (Homma and Yoshida, 2025, Equation 2):

$$\rho_j \in [L(p_{j,1}, p_{j,2}), U(p_{j,1}, p_{j,2})] \subseteq [-1, 1]$$

where:

$$L(p_{j,1}, p_{j,2}) = \max\left\{-\sqrt{\frac{p_{j,1}p_{j,2}}{(1-p_{j,1})(1-p_{j,2})}}, -\sqrt{\frac{(1-p_{j,1})(1-p_{j,2})}{p_{j,1}p_{j,2}}}\right\}$$

$$U(p_{j,1}, p_{j,2}) = \min\left\{\sqrt{\frac{p_{j,1}(1-p_{j,2})}{p_{j,2}(1-p_{j,1})}}, \sqrt{\frac{p_{j,2}(1-p_{j,1})}{p_{j,1}(1-p_{j,2})}}\right\}$$

If $p_{j,1} = p_{j,2}$, then $U(p_{j,1}, p_{j,2}) = 1$, whereas $L(p_{j,1}, p_{j,2}) = -1$ for $p_{j,1} + p_{j,2} = 1$.

These bounds can be computed using the `corrbound2Binary()` function in this package.

### Hypothesis Testing

We test superiority of treatment over control for both endpoints. The **endpoints** for testing are the risk differences:

**For endpoint** $k$:
$${\rm H}_{0k}: p_{1,k} - p_{2,k} \leq 0 \text{ vs. } {\rm H}_{1k}: p_{1,k} - p_{2,k} > 0$$

**For co-primary endpoints** (intersection-union test):

**Null hypothesis**: ${\rm H}_{0} = {\rm H}_{01} \cup {\rm H}_{02}$ (at least one null is true)

**Alternative hypothesis**: ${\rm H}_{1} = {\rm H}_{11} \cap {\rm H}_{12}$ (both alternatives are true)

**Decision rule**: Reject ${\rm H}_{0}$ at level $\alpha$ if and only if **both** ${\rm H}_{01}$ and ${\rm H}_{02}$ are rejected at level $\alpha$.

### Test Statistics

Several test statistics are available for binary endpoints. We focus on **asymptotic normal approximation** methods following Sozu et al. (2010).

#### Method 1: Standard Normal Approximation (AN)

For endpoint k, the test statistic without continuity correction is **(Equation 3 in Sozu et al., 2010)**:

$$Z_k = \frac{\hat{p}_{1,k} - \hat{p}_{2,k}}{se_{k0}}$$

where:
$$se_{k0} = \sqrt{\left(\frac{1}{n_1} + \frac{1}{n_2}\right) \bar{p}_k(1 - \bar{p}_k)}$$

and $\bar{p}_k = \frac{n_1 \hat{p}_{1,k} + n_2 \hat{p}_{2,k}}{n_1 + n_2}$ is the **pooled proportion** under the null hypothesis.

Under ${\rm H}_{0k}$, $Z_{k}$ asymptotically follows ${\rm N}(0, 1)$.

**Power formula** (Equation 4 in Sozu et al., 2010):

Under ${\rm H}_{1k}$ with true probabilities $p_{1,k}$ and $p_{2,k}$, the power for a single endpoint is:

$${\rm Power}_k = \Phi\left(\frac{\delta_k - se_{k0} z_{1-\alpha}}{se_k}\right)$$

where:

- $\delta_k = p_{1,k} - p_{2,k}$ is the true risk difference
- $se_k = \sqrt{\frac{v_{1,k}}{n_1} + \frac{v_{2,k}}{n_2}}$ with $v_{j,k} = p_{j,k}(1 - p_{j,k})$
- $z_{1-\alpha}$ is the $(1-\alpha)$ quantile of the standard normal distribution

#### Method 2: Normal Approximation with Continuity Correction (ANc)

To improve finite-sample performance, Yates's continuity correction is applied **(Equation 5 in Sozu et al., 2010)**:

$$Z_k = \frac{\hat{p}_{1,k} - \hat{p}_{2,k} - c}{se_{k0}}$$

where:
$$c = \frac{1}{2}\left(\frac{1}{n_1} + \frac{1}{n_2}\right)$$

**Power formula**:
$${\rm Power}_k = \Phi\left(\frac{\delta_k - se_{k0} z_{1-\alpha} - c}{se_k}\right)$$

#### Method 3: Arcsine Transformation (AS)

The arcsine-square-root transformation stabilizes variance **(Equation 6 in Sozu et al., 2010)**:

$$Z_k = \frac{\arcsin(\sqrt{\hat{p}_{1,k}}) - \arcsin(\sqrt{\hat{p}_{2,k}})}{se}$$

where:
$$se = \frac{1}{2}\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$

**Power formula**:

Let $\delta_k^{\rm AS} = \arcsin(\sqrt{p_{1k}}) - \arcsin(\sqrt{p_{2k}})$, then:

$${\rm Power}_k = \Phi\left(\frac{\delta_k^{\rm AS}}{se} - z_{1-\alpha}\right)$$

#### Method 4: Arcsine Transformation with Continuity Correction (ASc)

Walters' continuity correction for the arcsine method **(Equation 7 in Sozu et al., 2010)**:

$$Z_k = \frac{\arcsin(\sqrt{\hat{p}_{1,k} + c_1}) - \arcsin(\sqrt{\hat{p}_{2,k} + c_2})}{se}$$

where:
$$c_1 = -\frac{1}{2n_1}, \quad c_2 = \frac{1}{2n_2}$$

$$se = \frac{1}{2}\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$

**Power formula**:

Let:

- $\delta_k^{ASc} = \arcsin(\sqrt{p_{1,k} + c_1}) - \arcsin(\sqrt{p_{2,k} + c_2})$
- $v_{j,k}^c = (p_{j,k} + c_j)(1 - p_{j,k} - c_j)$
- $se_k = \sqrt{\frac{v_{1,k}}{4n_1 v_{1,k}^c} + \frac{v_{2,k}}{4n_2 v_{2,k}^c}}$

Then:
$${\rm Power}_k = \Phi\left(\frac{\delta_k^{ASc} - se \cdot z_{1-\alpha}}{se_k}\right)$$

### Joint Distribution and Correlation

Under ${\rm H}_{1}$, $(Z_{1}, Z_{2})$ asymptotically follows a bivariate normal distribution:

$$\begin{pmatrix} Z_1 \\ Z_2 \end{pmatrix} \sim {\rm BN}\left(\begin{pmatrix} \mu_1 \\ \mu_2 \end{pmatrix}, \begin{pmatrix} 1 & \gamma \\ \gamma & 1 \end{pmatrix}\right)$$

The correlation $\gamma$ between test statistics depends on $\rho_{1}$, $\rho_{2}$, and the marginal probabilities. For details on the derivation and approximation formulas, see Sozu et al. (2010) and Homma and Yoshida (2025). This approximation is implemented in the package functions.

### Power Calculation

The overall power (probability of rejecting both null hypotheses) is:

$$1 - \beta = {\rm P}(Z_1 > z_{1-\alpha} \text{ and } Z_2 > z_{1-\alpha} \mid {\rm H}_1)$$

Using the bivariate normal distribution:

$$1 - \beta = \Phi_2(-z_{1-\alpha} + \mu_1, -z_{1-\alpha} + \mu_2 \mid \gamma)$$

where $\Phi_{2}(a, b \mid \gamma)$ is the CDF of the standard bivariate normal distribution with correlation $\gamma$.

### Sample Size Calculation

For target power $1 - \beta$ and balanced design ($n_{1} = n_{2} = n$), we solve the power formula shown above numerically for $n_{2}$. Then, the total sample size is obtained as $N = (1+r)n_{2}$.

## Replicating Sozu et al. (2010) Table III

Table III from Sozu et al. (2010) shows sample sizes for various probability combinations and correlations using the AN, ANc, AS, and ASc test methods. Note that exact methods based on Fisher's exact test (Homma and Yoshida, 2025) can provide more accurate sample sizes for small to medium sample sizes, but are not included in this table as they were not available in Sozu et al. (2010).

The notation used in the function is: `p11` = $p_{1,1}$, `p12` = $p_{1,2}$, `p21` = $p_{2,1}$, `p22` = $p_{2,2}$, where the first subscript denotes the group (1 = treatment, 2 = control) and the second subscript denotes the endpoint (1 or 2).

```{r table3_sozu}
# Recreate Sozu et al. (2010) Table III
library(dplyr)
library(tidyr)
library(readr)

param_grid_bin_ss <- tibble(
  p11 = c(0.70, 0.87, 0.90, 0.95), 
  p12 = c(0.70, 0.70, 0.90, 0.95),
  p21 = c(0.50, 0.70, 0.70, 0.90),
  p22 = c(0.50, 0.50, 0.70, 0.90)
)

result_bin_ss <- do.call(
  bind_rows,
  lapply(c("AN", "ANc", "AS", "ASc"), function(test) {
    do.call(
      bind_rows,
      design_table(
        param_grid = param_grid_bin_ss,
        rho_values = c(-0.3, 0, 0.3, 0.5, 0.8),
        r = 1,
        alpha = 0.025,
        beta = 0.2,
        endpoint_type = "binary",
        Test = test
      ) %>% 
        mutate(Test = test)
    )
  })
) %>% 
  mutate_at(vars(starts_with("rho_")), ~ . / 2) %>% 
  pivot_longer(
    cols = starts_with("rho_"),
    names_to = "rho",
    values_to = "N",
    names_transform = list(rho = parse_number)
  ) %>% 
  pivot_wider(names_from = Test,  values_from = N) %>% 
  drop_na(AN, ANc, AS, ASc) %>% 
  as.data.frame()

kable(result_bin_ss,
      caption = "Table III: Sample Size per Group (n) for Two Co-Primary Binary Endpoints (α = 0.025, 1-β = 0.80)^a,b^",
      digits = 0,
      col.names = c("p₁,₁", "p₁,₂", "p₂,₁", "p₂,₂", "ρ", "AN", "ANc", "AS", "ASc"))
```

^a^ AN: Asymptotic normal test without continuity correction; ANc: Asymptotic normal test with continuity correction; AS: Arcsine transformation without continuity correction; ASc: Arcsine transformation with continuity correction.

^b^ Some values may differ by 1-2 subjects from Sozu et al. (2010) Table III due to numerical differences in computing the bivariate normal cumulative distribution function between SAS and R implementations. Power calculations at the reported sample sizes confirm the accuracy of the values presented here.

### Key Findings

**Effect of correlation**:

- **Positive correlation** ($\rho > 0$) reduces required sample size
- **Negative correlation** ($\rho < 0$) increases required sample size
- **Zero correlation** ($\rho = 0$) provides intermediate sample size

**Test method comparison**:

- **AN**: Generally smallest sample size, most efficient
- **ANc**: Slightly larger than AN due to continuity correction
- **AS**: Similar to AN for moderate probabilities
- **ASc**: Largest sample size, most conservative

## Basic Usage Examples

### Example 1: Equal Effect Sizes

Calculate sample size when both endpoints have equal effect sizes:

```{r example1}
# Both endpoints: 70% vs 50% (20% difference)
# p_{1,1} = p_{1,2} = 0.7 (treatment group)
# p_{2,1} = p_{2,2} = 0.5 (control group)
ss_equal <- ss2BinaryApprox(
  p11 = 0.7, p12 = 0.7,  # Treatment group
  p21 = 0.5, p22 = 0.5,  # Control group
  rho1 = 0.5,            # Correlation in treatment group
  rho2 = 0.5,            # Correlation in control group
  r = 1,                 # Balanced allocation
  alpha = 0.025,
  beta = 0.2,
  Test = "AN"
)

print(ss_equal)
```

**Note**: In the function, `p11` corresponds to $p_{1,1}$ (success probability for endpoint 1 in treatment group), `p12` to $p_{1,2}$, `p21` to $p_{2,1}$, and `p22` to $p_{2,2}$.

### Example 2: Unequal Effect Sizes

When effect sizes differ, the endpoint with smaller effect dominates:

```{r example2}
# Endpoint 1: 75% vs 65% (10% difference)
# Endpoint 2: 80% vs 60% (20% difference)
ss_unequal <- ss2BinaryApprox(
  p11 = 0.75, p12 = 0.80,
  p21 = 0.65, p22 = 0.60,
  rho1 = 0.3, rho2 = 0.3,
  r = 1,
  alpha = 0.025,
  beta = 0.2,
  Test = "AN"
)

print(ss_unequal)

# Compare with individual endpoint sample sizes
ss_ep1 <- ss1BinaryApprox(p1 = 0.75, p2 = 0.65, r = 1, 
                          alpha = 0.025, beta = 0.2, Test = "AN")
ss_ep2 <- ss1BinaryApprox(p1 = 0.80, p2 = 0.60, r = 1,
                          alpha = 0.025, beta = 0.2, Test = "AN")

cat("Single endpoint 1 sample size:", ss_ep1$n2, "\n")
cat("Single endpoint 2 sample size:", ss_ep2$n2, "\n")
cat("Co-primary sample size:", ss_unequal$n2, "\n")
```

The co-primary design requires approximately the same sample size as the endpoint with smaller effect size.

### Example 3: Effect of Correlation

Demonstrate how correlation affects sample size:

```{r correlation_effect}
# Fixed effect sizes: p_{1,1} = p_{1,2} = 0.7, p_{2,1} = p_{2,2} = 0.5
p11 <- 0.7
p21 <- 0.5
p12 <- 0.7
p22 <- 0.5

# Range of correlations
rho_values <- c(-0.3, 0, 0.3, 0.5, 0.8)

ss_by_rho <- sapply(rho_values, function(rho) {
  ss <- ss2BinaryApprox(
    p11 = p11, p12 = p12,
    p21 = p21, p22 = p22,
    rho1 = rho, rho2 = rho,
    r = 1,
    alpha = 0.025,
    beta = 0.2,
    Test = "AN"
  )
  ss$n2
})

result_df <- data.frame(
  rho = rho_values,
  n_per_group = ss_by_rho,
  N_total = ss_by_rho * 2
)

kable(result_df,
      caption = "Sample Size vs Correlation (p₁,₁ = p₁,₂ = 0.7, p₂,₁ = p₂,₂ = 0.5)",
      col.names = c("ρ", "n per group", "N total"))

# Plot
plot(rho_values, ss_by_rho, 
     type = "b", pch = 19,
     xlab = "Correlation (ρ)", 
     ylab = "Sample size per group (n)",
     main = "Effect of Correlation on Sample Size",
     ylim = c(min(ss_by_rho) * 0.9, max(ss_by_rho) * 1.1))
grid()
```

**Interpretation**: Higher positive correlation substantially reduces required sample size.

### Example 4: Unbalanced Allocation

Calculate sample size with unequal group allocation:

```{r unbalanced}
# 2:1 allocation (treatment:control)
ss_unbalanced <- ss2BinaryApprox(
  p11 = 0.7, p12 = 0.7,
  p21 = 0.5, p22 = 0.5,
  rho1 = 0.5, rho2 = 0.5,
  r = 2,  # 2:1 allocation
  alpha = 0.025,
  beta = 0.2,
  Test = "AN"
)

print(ss_unbalanced)
cat("Total sample size:", ss_unbalanced$N, "\n")
```

## Power Verification

Verify that calculated sample sizes achieve target power:

```{r power_verification}
# Use result from Example 1
power_result <- power2BinaryApprox(
  n1 = ss_equal$n1,
  n2 = ss_equal$n2,
  p11 = 0.7, p12 = 0.7,
  p21 = 0.5, p22 = 0.5,
  rho1 = 0.5, rho2 = 0.5,
  alpha = 0.025,
  Test = "AN"
)

cat("Target power: 0.80\n")
cat("Achieved power (Endpoint 1):", round(power_result$power1, 4), "\n")
cat("Achieved power (Endpoint 2):", round(power_result$power2, 4), "\n")
cat("Achieved power (Co-primary):", round(power_result$powerCoprimary, 4), "\n")
```

## Comparison of Test Methods

Compare the four asymptotic test methods:

```{r method_comparison}
# Fixed design parameters
p11 <- 0.80
p12 <- 0.70
p21 <- 0.55
p22 <- 0.45
rho <- 0.7

# Calculate for each method
methods <- c("AN", "ANc", "AS", "ASc")
comparison <- lapply(methods, function(method) {
    ss <- ss2BinaryApprox(
        p11 = p11, p12 = p12,
        p21 = p21, p22 = p22,
        rho1 = rho, rho2 = rho,
        r = 1,
        alpha = 0.025,
        beta = 0.2,
        Test = method
    )
    
    data.frame(
        Method = method,
        n_per_group = ss$n2,
        N_total = ss$N
    )
})

comparison_table <- bind_rows(comparison)

kable(comparison_table,
      caption = "Comparison of Test Methods (p₁,₁ = p₁,₂ = 0.7, p₂,₁ = p₂,₂ = 0.5, ρ = 0.5)")
```


## Practical Recommendations

### Design Considerations

1. **Estimate correlation**: Use pilot data or historical information; be conservative if uncertain

2. **Consider asymmetric effects**: Use actual effect sizes rather than assuming equality

3. **Balanced allocation**: Generally more efficient unless practical constraints require otherwise

4. **Test method selection**: AN is most common; use continuity correction for added conservatism

5. **Sample size sensitivity**: Calculate for range of plausible correlations and effect sizes

### When to Use Exact Methods Instead

Use exact methods (`ss2BinaryExact`) based on Homma and Yoshida (2025) when:

- Small to medium sample sizes ($N < 200$)
- Extreme probabilities ($p < 0.1$ or $p > 0.9$)
- Strict Type I error control required
- Regulatory preference for exact tests

Exact methods using Fisher's exact test can provide more accurate sample sizes and better Type I error control in these situations.

### Asymptotic Validity

These asymptotic methods are appropriate when:

- $N > 200$
- $0.1 < p < 0.9$
- Computational efficiency is important

## References

Homma, G., & Yoshida, T. (2025). Exact power and sample size in clinical trials with two co-primary binary endpoints. *Statistical Methods in Medical Research*, 34(1), 1-19.

Sozu, T., Sugimoto, T., & Hamasaki, T. (2010). Sample size determination in clinical trials with multiple co-primary binary endpoints. *Statistics in Medicine*, 29(21), 2169-2179.
